---
title: "Class 19: Pertussis Mini Project"
author: "Shawn Xiao (PID: A18276668)"
format: pdf
toc: True
---

## Background

Pertussis is a bacterial lung infection also known as whooping cough. 
Let's begin by examining CDC reported case numbers in the US. 

```{r}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L,2024L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063,22538)
       )
```

Plor of cases per year for pertussis in the US

```{r}
library(ggplot2)
```

```{r}
ggplot(cdc)+
  aes(year,
      cases)+
  geom_line()+
  geom_point()
```

Add some major milestone timeplints to our plot: 

```{r}
ggplot(cdc)+
  aes(year,
      cases)+
  geom_line()+
  geom_point()+
  geom_vline(xintercept = 1947, col = "blue")+
  geom_vline(xintercept = 1996, col = "darkred")+
  geom_vline(xintercept = 2020, col = "orange")
```

The full introduction of mandatory wP (whole-cell) Pertussis imunization in the 1940s lead to a dramatic reduction in case numbers(from over 200,000 to 100s). 

The switch to the aP (nwer acellular) formalization lead to a slight increase in case numbers because of the anti-vaccine movement and the effectiveness of vaccine. 

The COVID pendanmic in 2020 lead to a decrease in case numbers since people have more public health awearness. 

## The CMI-PB Project

The CMI-PB project aims to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

They make their data available via JSON format API endpoints- basically the database tables in a key: value type format like "infancy_vac":"wP". To read this we can use the `read_json()` function from the **jsonlite** package. 

```{r}
library(jsonlite)

subject <- read_json(path="https://www.cmi-pb.org/api/v5_1/subject",
                     simplifyVector = T)
head(subject)
```

> Q. How many subject/individuals are in this dataset? 

```{r}
nrow(subject)
```

There are `r nrow(subject)` subject/individuals in this dataset. 

> Q. How many wP and aP subjects are there

```{r}
table(subject$infancy_vac)
```

There are 87 aP subjects. 

> Q. What is the breakdown by "biologicak_sex" and "race"?

```{r}
table(subject$biological_sex)
table(subject$race)
```

```{r}
table(subject$race, subject$biological_sex)
```

This breakdown is not particularly representative of the US population - this is a serious cavet for this study. However, it is still the largest sample of its type every assembled. 

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen",simplifyVector = T)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer",simplifyVector = T)
```

```{r}
head(specimen)
```

We need to "join" or link these tables with the `subject` table so we can begin to analyze this data and know who a given Ab sample was collected for and when. 

```{r,message=FALSE}
library(dplyr)

meta <- inner_join(subject,specimen)
head(meta)
```

Now let's join the `ab_titer` table with our `meta   table so we have all information about a given measurement. 

```{r}
ab_data <- inner_join(ab_titer,meta)
head(ab_data)
```

> Q. how many Ab measurements do we have

We have `r nrow(ab_data)` measurements. 

> Q. How many different isotypes (types of Ab) are in the dataset

```{r}
unique(ab_data$isotype)
```

> Q How many antigens are there

```{r}
unique(ab_data$antigen)
```

```{r}
ggplot(ab_data)+
  aes(MFI,antigen)+
  geom_boxplot()
```


## Examine IgG Ab titer levels

IgG is crucial for long-term imunity and responding to bacterial  and viral infections. 

```{r}
igg <- ab_data|>
  filter(isotype=="IgG")
```

```{r}
ggplot(igg)+
  aes(MFI_normalised,antigen)+
  geom_boxplot()
```

We can "facet" our plot by wP vs. aP

```{r}
ggplot(igg)+
  aes(MFI_normalised,antigen, col=infancy_vac)+
  geom_boxplot()+
  facet_wrap(~infancy_vac)
```

```{r}
ggplot(igg)+
  aes(MFI_normalised,antigen, col=infancy_vac)+
  geom_boxplot()
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```


More advanced analysis digging into individual antigen responses over time: 

```{r}
filter(igg, antigen=="PT",dataset == "2021_dataset") |>
  ggplot() +
  aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point()+
  geom_line()+
  geom_vline(xintercept=14, linetype="dashed")+
  theme_classic()
```

The plot shows the time course of Pertussis toxin (PT) antibody responses for a large set of wP (teal) and aP (red) individuals. Levels peak at day 14 and are larger in magnitude for wP than aP individuals. 

There are lots of cool things to explore in this dataset and we need coding and biology knowledge to do it effectively. 

