---
title: "Class 8: Mini Project"
author: "Shawn Xiao (PID:A18276668)"
format: pdf
toc: True
---

## Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class. You’ll extend what you’ve learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses. This expands on our RNA-Seq analysis from last day.

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.

## Data Import

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
head(wisc.df)
```


Remove the Diagnosis column for subsequent analysis

```{r}
wisc.data <- wisc.df[,-1]
```

```{r}
diagnosis <- factor(wisc.df$diagnosis)
```

>Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```

There are `r nrow(wisc.data)` observations in the dataset. 

>Q2. How many of the observations have a malignant diagnosis?

```{r}
table(diagnosis)
```

212 observations have a malignant diagnosis. 

>Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean",colnames(wisc.data)))
```

There are 10 variables in the data that are suffixed with "_mean". 



## Principal Component Analysis (PCA)

The`prcomp()` function to do PCA has a `scale=F` default. In general sclae=T should always be used since certain variables can have a high mean and standard deviation that can skew the data. 

```{r}
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

```{r}
wisc.pr <- prcomp(wisc.data,scale=T)
summary(wisc.pr)
```

>Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

44.27% of the original variance is captured by PC1. 

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 principal components are required to describe at least 70% of the original variance. 

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 principal components are required to describe at least 90% of the original variance. 

### PCA score plot

The main plot created by the PCA scores is scatter plot. 

```{r}
library(ggplot2)

ggplot(wisc.pr$x)+
  aes(PC1,PC2, col=diagnosis)+
  geom_point()
```

```{r}
biplot(wisc.pr)
```

>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

The plot has four axis and is very hard to understand because every variable is labeled and the pattern of the graph is blocked. 

```{r}
plot(wisc.pr$x , col = diagnosis , 
     xlab = "PC1", ylab = "PC2")
```

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
ggplot(wisc.pr$x)+
  aes(PC1,PC3, col=diagnosis)+
  geom_point()
```

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
pve <- pr.var/sum(pr.var)
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

### Communicating PCA results

>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean",1]
```

The loading factor for concave.points_mean is `r wisc.pr$rotation["concave.points_mean",1]`. 

>Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

The minimum number of principal components required to explain 80% of th variance is 5. 

## Hierarchical Plot

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist,method="complete")
```

### Results of hierarchical clustering

>Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```

### Selecting number of clusters

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust,k=4)

table(wisc.hclust.clusters, diagnosis)
```

>Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

No

>Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
wisc.hclust.single<- hclust(data.dist,method="single")
wisc.hclust.average <- hclust(data.dist,method="average")
wisc.hclust.ward.D2 <- hclust(data.dist,method="ward.D2")
plot(wisc.hclust.single)
plot(wisc.hclust.average)
plot(wisc.hclust.ward.D2)
```

I like `"ward.D2"` the best because it gives a plot with the clearest branches. 

## Combining Method (PCA and Clustering)

Clustering the original data was not very productive. The PCA results looked promising. Here we combine these methods by clustering from our PCA results. Using PCA results to create clustering. 

```{r}
## take the first three PCs
dist.pc <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(dist.pc,method="ward.D2")
```

View the tree

```{r}
plot(wisc.pr.hclust)
abline(h=80, col="red", lty=2)
```

To get our clustering membership vector (i.e. our main clustering result) we "cut" the tree at a desired height. 

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

>Q15. How well does the newly created model with four clusters separate out the two diagnoses?

```{r}
table(grps, diagnosis)
```

The new model can correctly separate patients by diagnosis and by two groups from the cluster diagram, which indicates the patients' actual tumor type. 

>Q15. Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
wisc.km <- kmeans(wisc.data, centers= 2)
table(wisc.km$cluster, diagnosis)
table(wisc.hclust.clusters, diagnosis)
```

The `kmeans()` and `hclust` functions cannot separate the actual diagnosis as well as the PCA because they put data as discrete variables and sort them merely by distance whereas PCA views data as a continuous variable and is less sensitive to noises for large dimension data. 

Scatter plot

```{r}
## colored by grps
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
## Colored by diagnosis
plot(wisc.pr$x[,1:2], col=diagnosis)
```

>Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

Sensitivity: TP/(TP+FN)
Specificity: TN/(TN+FN)

```{r}
179/(179+33)
333/(333+33)
```

The PCA method resulted in the best specificity and sensitivity. 



## Prediction

We can use our PCA model to predict

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
g <- as.factor(grps)
levels(g)

g <- relevel(g,2)
levels(g)

plot(wisc.pr$x[,1:2], col=g)
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

>Q18. Which of these new patients should we prioritize for follow up based on your results?

Patient 2

