---
title: "Class 5: Data Viz With gglot"
author: "Shawn Xiao(PID: A18276668)"
format: pdf
---

Today we are exploring the **ggplot** package and how to make nice figures in R. 

There are lots of ways to make figures and plots in R. These include: 

- so called "base" R
- and add on packages like **ggplot**

Here is a simple "base" R plot. 

```{r}
head(cars)
```

We can simply pass this to the `plot()` function. 

```{r}
plot(cars)
```

> Key-point: Base R is quick but not so nice looking in some folks' eyes

Let's see how we can plot this with **ggplot**

1st I need to install this add-on package. For this we use the `install.packages()` function - *WE DO THIS IN THE CONSOLE, NOT our report*. This is a one time only deal. 

2nd We need to load the package with the `library()` function every time we want to use it. 

```{r}
library(ggplot2)
```

Every ggplot is composed of at least 3 layers: 

- **data** (i.e. a data.frame with the things you want to plot)
- aesthetics **aes()** that map the columns of data to your plot features (i.e. aesthetics)
- geoms like **geom_point()** that srt how the plot appears. 

```{r}
ggplot(cars) +
  aes(x=speed, y=dist)+
  geom_point()
```

>Key-point: For simple "canned" graphs base R is quicker but as things get more custom and elaborate then ggplot wins out...

Let's add more layers to out ggplot

Add a line showing the relationship between x and y
Add title of the graph
Add titles for x and y axis
Change theme

```{r}
ggplot(cars) +
  aes(x=speed, y=dist)+
  geom_point()+
  geom_smooth(method="lm", se=FALSE)+
  labs(title="Silly Plot of Speed vs Stoping Distnace")+
  xlab("Speed(MPH)")+
  ylab("Distance(ft)")+
  theme_bw()
```


## Going Further

Read gene expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```


> Q1. How many genes are in this dataset?

```{r}
nrow(genes)
```

> Q2. How many "up" regulated genes are there?

```{r}
sum(genes$State=="up")
```

A useful function for counting up occurances of things in a vector is the `table()` function

```{r}
table(genes$State)
```


```{r}
p <- ggplot(genes)+
  aes(x=Condition1,y=Condition2)+
  geom_point()+
  aes(col=State)

p +
   scale_color_manual(values=c("darkred","darkgrey","navy"))+
  labs(title="Expression changes upon drug treatment",
       x="control (no drug)",
       y="Treatment (with drug)")+
  theme_bw()
```

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder,3)
```

```{r}
tail(gapminder,3)
```

>Q4. How many countries are in this data set?

```{r}
length(table(gapminder$country))
```

>Q4. How many continents are in this data set?

```{r}
unique(gapminder$continent)
length(table(gapminder$continent))
```

```{r}
ggplot(gapminder)+
  aes(x=gdpPercap, y=lifeExp,
      col=continent)+
  geom_point()
```

```{r}
ggplot(gapminder)+
  aes(x=gdpPercap, y=lifeExp,
      col=continent,
      label=country)+
  geom_point()+
  geom_text()
```

I can you **ggrepel** package to make more sensible labels here. 

```{r}
library(ggrepel)
```


I want seperate pannel per continent

```{r}
ggplot(gapminder)+
  aes(x=gdpPercap, y=lifeExp,
      col=continent,
      label=country)+
  geom_point()+
  geom_text_repel()+
  facet_wrap(~continent)
```

## Summary
ggplot2 makes it easier to create publication-quality, beautiful figures and allows you to build plots layer by layer using a consistent grammar. It is more concise for complex plots, offers sensible defaults, and handles legends and layout automatically. Base R plots are faster for simple, exploratory graphs but are harder to customize and refine for polished results. ggplot2 is preferred for finalized, professional visualizations, while base R is often used for quick, preliminary plots [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).