# ---
title: "Class06: R functions"
author: "Shawn Xiao (PID:A18276668)"
format: gfm
---

All functions in R have at least 3 things:

-   A **name**, we pick this and use it to call the function.
-   Input **arguments**, there an be multiple comma seperated inputs to the function.
-   The **body**, lines of R code that do the work of the function.

Our first wee function:

```{r}
add <- function(x){
  x+1
}
```

Let's test our function

```{r}
add(10)
```

```{r}
add12 <- function(x,y=1){
  x+y
}
```

```{r}
add12(10)
```

```{r}
add12(10,10)
```



## A second function

Let's try something more interesting. Make a sequence generation tool

The `sample()` function could be useful here

```{r}
sample(1:10, size=3)
```

```{r}
sample(c("A","T","G","C"),size=5,replace=TRUE)
```

Turn this snipet into a function that returns a user sepcified length DNA squence, let's call it `generate_dna`
```{r}
generate_dna <- function(x=10){
  v <- sample(c("A","T","G","C"),size=x,replace=TRUE)
  cat("Well Done You!\n")
  return(v)
}
```

```{r}
generate_dna(5)
```

```{r}
s <- generate_dna(15)
s
```

I want the option to return a single element vector like "GGAGTAC"

```{r}
paste(s,collapse="")
```

```{r}
generate_dna <- function(x=10, fasta=FALSE){
  v <- sample(c("A","T","G","C"),size=x,replace=TRUE)
  
  cat("Well Done You!\n")
  
  # Make a single element vector
  s <- paste(v,collapse="")
  
  # Adding fasta logic
  if(fasta){
    return(s)
  } else{
    return(v)
  }
}
```


```{r}
generate_dna()
```

```{r}
generate_dna(10, fasta=T)
```

```{r}
generate_prot <- function(ami=10, fasta=TRUE){
  a <- c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")
  p <- sample(a,size=ami,replace=TRUE)
  
  
  # Make a single element vector
  t <- paste(p,collapse="")
  
  # Adding fasta logic
  if(fasta){
    return(t)
  } else{
    return(p)
  }
}
```


```{r}
generate_prot(20)
```



>Q. Generate random protein sequencies between lengths 5 and 12 amino acids. 

One approach is to do this by brute force calling our function for each length 5 to 12

Another approach is to write a `for()` loop to itterate over the input valued 5 to 12

A very useful third R specific apprach is to use the `sapply()` function. 

```{r}
seq_length <- 5:12
for(i in seq_length){
  cat(">",i,"\n")
  cat(generate_prot(i))
  cat("\n")
}
```


```{r}
sapply(5:12,generate_prot)
```


> **Key Point**: Writing functions in R is doable but not the easiest thing in the world. Starting with a working snippet of code and then using LLM tools to improve and generate addative functions to make it more useful. 