---
title: "Class 7: Machine Learning 1"
author: "Shawn Xiao (PID:A18276668)"
format: pdf
---

Today we will explore some fundamental machine learning methods including clustering and dimensionality reduction. 

## K-means clustering 

To see how this works let's first makeup some data to cluster where we know what the answer should be. We can use the `rnorm()`funciton to help: 

```{r}
hist(rnorm(500,mean=5))
```


```{r}
x <- c(rnorm(30,mean=-3),rnorm(30,mean=3))
y <- rev(x)
```

```{r}
x <- cbind(x,y)
plot(x)
```
The function for K-means clustering in "base" R is `kmeans()`

```{r}
k <- kmeans(x,centers=2)
k
```


To get at the results of the returned list object we can use the `$` syntax

>Q. How many points are in each cluster?

```{r}
k$size
```

>Q. What ‘component’ of your result object details
 - cluster size?
 - cluster assignment/membership?
 - cluster center?
 
```{r}
k$cluster
k$centers
```


>Q. Make a clustering results figure of the data colored by cluster membership and show cluster centers. 

```{r}
plot(x,col=c("red","blue"))
```

```{r}
plot(x,col=k$cluster,pch=16)
points(k$centers,col="navy",pch=15,cex=2)
```

K-means clustering is very popular as it is very fast and relatively stright forward: it takes numeric data as input and returns the cluster membership vector. 

The "issue" is we tell `kmeans()` how many clusters we want!

>Q. Run kmeans() again and cluster into 4 clusters and plot the results like we did before? 

```{r}
k4 <- kmeans(x,centers=4)
plot(x,col=k4$cluster,pch=16)
points(k4$centers,col="navy",pch=15,cex=2)
```

Scree plot to pick k `centers` value

brute force or `for()` loop


```{r}
n <- NULL
for(i in 1:5){
  
  n <- c(n, kmeans(x,centers=i)$tot.withinss)
  
}
plot(n,typ="b")
```


## Hierarchical Clustering

The main "base" R function for Hierarchical Clustering is called`hclust()`. Here we can't just input our data we need to fitst caculate a distance matrix (`dist()`) for our data and use this as input to `hclust()`

```{r}
d <- dist(x)
hc <- hclust(d)
hc
```


The is a plot method for hclust results 

```{r}
plot(hc)
abline(h=8,col="red")
```


To get out cluster "membership" vector(i.e. our main clustering result) we can "cut" the tree at a given height or at a height that yeilds a given "k" groups
```{r}
cutree(hc,h=8)
```

```{r}
grps <- cutree(hc, k=2)
```


>Q. Plot the data with our hclust result coloring

```{r}
plot(x, col=grps, pch=16)
```


# Principle Componenet Analysis (PCA)

## PCA for UK food data

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
head(x)
```


```{r}
rownames(x) <- x[,1]
x <- x[,-1]
x
```


```{r}
x <- read.csv(url,row.names = 1)
```

Some base functions

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

```{r}
barplot(as.matrix(x), col=rainbow(nrow(x)))
```

There is one plot that can be useful for small datasets: 

```{r}
pairs(x, col=rainbow(10), pch=16)
```

>Main point: It can be difficult to spot major trends and patterns even in relatively small multivariate datasets (here we only have 17 dimensions, typically we have 1000s)

## PCA to the resecue

The main function in "base" R for PCA is called `prcomp()`

```{r}
# I will take transpose of the food data `t(x)`
pca <- prcomp( t(x) )
summary(pca)
```

```{r}
cols <- c("orange","red","blue","darkgreen")
plot(pca$x[,1], pca$x[,2],col=cols,pch=16)
```

```{r}
library(ggplot2)
```

```{r}
ggplot(pca$x)+
  aes(PC1,PC2)+
  geom_point(col=cols)
```

```{r}
ggplot(pca$rotation)+
  aes(PC1, rownames(pca$rotation))+
  geom_col()
```


PCA looks super useful and we will come back to describe this further next day. 

