---
title: "Class 10: Halloween Mini-Project"
author: "Shawn Xiao (A18276668)"
format: pdf
toc: True
---

## Background

As it is nearly Halloween and the half way point in the quarter, let's do a mini project that help us figure out the best candy!

Our data come drom the 538 website and is available as a CSV file. 

## Data Import

```{r,message=FALSE}
candy <- read.csv(file="candy-data.txt",row.names=1)
flextable::flextable(head(candy,10))
```


>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

There are `r nrow(candy)` candy types in this dataset. 

>Q2. How many fruity candy types are in the dataset?

```{r,message=FALSE}
library(tidyverse)
sum(candy$fruity)
```

There are 38 fruity candy types in the dataset. 

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r, message=FALSE}
candy%>%
  filter(rownames(candy)=="Kit Kat")%>%
  select(winpercent)
```

My favorate candy is Kit Kat and its win rate is 76.77%. 

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

The winpercent value for Kit Kat is 76.77%

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy%>%
  filter(rownames(candy)=="Tootsie Roll Snack Bars")%>%
  select(winpercent)
```

The winpercent for Toosie Roll Snack Bars is 49.65%. 


## Quick Overview of Data

```{r}
skimr::skim(candy)
```


>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent is on a 0-100 scale and the rest are on 0-1 scale

>Q7. What do you think a zero and one represent for the candy$chocolate column?

The candy does not contain chocolate. 

>Q8. Plot a histogram of winpercent values

```{r}
ggplot(candy)+ 
  aes(winpercent)+
  geom_histogram(bins=20)
```

>Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy)+ 
  aes(winpercent)+
  geom_density()
```


The distribution is not symmetrical. 

>Q10. Is the center of the distribution above or below 50%?


```{r}
summary(candy$winpercent)
```

The center od the distribution is below 50%. 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy 
# 2. Find their winpercent 
# 3. Calculate the mean
# 4-6 Repeat with fruit candy

choc.win <- candy%>%
  filter(chocolate==1)%>%
  select(winpercent)

choc.mean <- mean(choc.win$winpercent)

friut.win <- candy%>%
  filter(fruity==1)%>%
  select(winpercent)

fruit.mean <- mean(friut.win$winpercent)
```


>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win,friut.win)
```


## Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

```{r}
candy |>
  arrange(winpercent)|>
  head(5)
```

The five least liked candy types are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy|>
  arrange(-winpercent)|>
  head(5)
```

The top five most popular candies are Reese's Peanut Butter Cup, Resse's Miniatures, Twix, Kit Kat, and Snickers. 

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy)+
  aes(winpercent,rownames(candy))+
  geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy)+
  aes(x=winpercent,
      y=reorder(rownames(candy),winpercent))+
  geom_col()
```

Add some color based on the "type of candy"

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] <-  "chocolate"
my_cols[as.logical(candy$bar)] <-  "orange"
my_cols[as.logical(candy$fruity)] <-  "brown"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```


## Winpercent and Pricepercent

A plot with both variables/columns winpercent and pricepercent

```{r}
my_cols[as.logical(candy$fruity)] <-  "brown"

ggplot(candy)+
  aes(winpercent,pricepercent,
      label=rownames(candy))+
  geom_point(col=my_cols)+
  geom_text()
```

```{r,message=FALSE}
library(ggrepel)

ggplot(candy)+
  aes(winpercent,pricepercent,
      label=rownames(candy))+
  geom_point(col=my_cols)+
  geom_text_repel(max.overlaps = 6)
```


>Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is sixlets. 

>Q18. What is the best ranked fruity candy?

The best ranked fruity candy is starbust. 

## Taking a look at pricepercent

```{r}

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures is ranked highest for winpercent and lowest for pricepercent. 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The top 5 most expensive candy that are least popular are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate.



## Exploring the correlation structure

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r,message=FALSE}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and Chocolate are anti-correlated. 

> Q23. Similarly, what two variables are most positively correlated?

Choloclate and Bar are most positively correlated. 

## Principal Component Analysis

The function to use is called `prcomp()` with an optional `scale=T/F` argument. 

```{r}
pca <- prcomp(candy,scale=T)
summary(pca)
```

Our main PCA result figure

```{r}
ggplot(pca$x)+
  aes(PC1,PC2,
      label=rownames(pca$x))+
  geom_point(col=my_cols)+
  geom_text_repel(col=my_cols, max.overlaps = 7)
```

We should also examine the "variable loading" or contributions of the original variables to the new PCs. 

```{r}
ggplot(pca$rotation)+
  aes(PC1,rownames(pca$rotation))+
  geom_col()
```

```{r}
p <- ggplot(pca$x)+
  aes(PC1,PC2,
      label=rownames(pca$x))+
  geom_point(col=my_cols)+
  geom_text_repel(col=my_cols, max.overlaps = 7)
p
```

Interactive plots that can be zoomed in and "brushed" over can be made with he **plotly** package. Its output is interactive and will not render to PDF. 

```{r}
library(plotly)
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, pluribus are picked up strongly by PC1 in the positive direction. This makes sense because candies with these three characteristics tend to correlate less with candies of other characteristics. 
